---
title: "Asynchronous Daemons with siphon and liteq"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Asynchronous Daemons with siphon and liteq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#."
)
```

## Introduction

In modern web applications built with frameworks like **Shiny** or **Plumber**, running long-running or resource-intensive computations directly in the main request handler blocks the user interface thread. This leads to high latency and a poor user experience.

A robust production architecture separates concerns:

1. **The Web Server (Shiny/Plumber)**: Quickly accepts user requests, writes jobs to a persistent queue (like `liteq`), and queries a database asynchronously to display results.
2. **The Worker Daemon**: A background R process that runs indefinitely, consumes tasks from the queue, processes them in parallel with controlled concurrency via `siphon`, and saves results back to the database.

This vignette demonstrates how to implement this background daemon pattern using `siphon` and `liteq`.

It builds on the core concepts - `pump()`, `pump_source()`, and `pump_drain()` - introduced in [Pull-Based Staged Pipelines with siphon](siphon.html) (`vignette("siphon", package = "siphon")`). The examples use the `liteq`, `mirai`, and `jsonlite` packages; install them to run the code below.

## Why `liteq`?

The `liteq` package provides a serverless, database-backed queue using SQLite. It is concurrent-safe, supports lock-free message consumption, and ensures that messages are either permanently acknowledged (`ack`) or returned to the queue on failure (`nack`), providing transaction-like reliability.

## Architecture

```{r daemon-diagram, echo = FALSE, fig.width = 7, fig.height = 2.2, fig.cap = "A web front end enqueues jobs; a siphon daemon consumes them, processes in parallel, and persists results."}
op <- par(mar = c(0, 0, 0, 0), xaxs = "i", yaxs = "i")
plot.new()
plot.window(xlim = c(0, 1), ylim = c(0, 1))

draw_box <- function(x, w, label, fill) {
  rect(x, 0.32, x + w, 0.78, col = fill, border = "#3C3C3C", lwd = 1.2)
  text(x + w / 2, 0.55, label, cex = 0.8)
}
draw_arrow <- function(x0, x1) {
  arrows(x0, 0.55, x1, 0.55, length = 0.08, lwd = 1.1, col = "#555555")
}

draw_box(0.02, 0.21, "Shiny / Plumber\npublish()", "#EEF3FB")
draw_box(0.30, 0.16, "liteq queue\n(SQLite)", "#FFF7DF")
draw_box(0.52, 0.22, "siphon daemon\npump_drain()", "#EEF6EA")
draw_box(0.81, 0.17, "Database\nack() + save", "#FBECEC")

draw_arrow(0.23, 0.30)
draw_arrow(0.46, 0.52)
draw_arrow(0.74, 0.81)

par(op)
```

## Step 1: A custom queue source

We wrap `liteq` consumption in `pump_source()`. Two design points are essential:

- **Keep the message on the main process.** A `liteq` message holds an open lock connection and must *not* be serialized to a parallel worker. We store each message in a main-process registry keyed by `id` and send only its (JSON) payload downstream. Workers compute on plain data; `ack()`/`nack()` happen later on the main process.
- **Release resources with `close_fn`.** `pump_run()` and `pump_drain()` call it automatically when the pipeline finishes.

A daemon runs forever (`done_fn = function() FALSE`); for the runnable demo in Step 2 we stop when the queue drains. The factory below supports both:

```{r}
library(siphon)

make_liteq_source <- function(queue, run_forever = TRUE) {
    registry <- new.env(parent = emptyenv())

    pump_source(
        pull_fn = function() {
            msg <- liteq::try_consume(queue)   # a liteq_message, or NULL if empty
            if (is.null(msg)) return(NULL)

            # Keep the live message on the main process so we can ack/nack it later
            assign(as.character(msg$id), msg, envir = registry)

            payload <- tryCatch(
                jsonlite::fromJSON(msg$message),
                error = function(e) e
            )
            list(
                id = msg$id,
                data = payload,                # only serializable data goes downstream
                ok = !inherits(payload, "error")
            )
        },
        item_commit_fn = function(id, data) {
            msg <- get(as.character(id), envir = registry, inherits = FALSE)
            liteq::ack(msg)
        },
        item_abort_fn = function(id, error = NULL, data = NULL) {
            msg <- get(as.character(id), envir = registry, inherits = FALSE)
            liteq::nack(msg)
        },
        item_release_fn = function(id) {
            key <- as.character(id)
            if (exists(key, envir = registry, inherits = FALSE)) {
                rm(list = key, envir = registry)
            }
        },
        done_fn = if (run_forever) function() FALSE else function() liteq::is_empty(queue),
        close_fn = function() {
            rm(list = ls(envir = registry), envir = registry)
            invisible(NULL)
        }
    )
}
```

## Step 2: A runnable end-to-end example

Let's exercise the source against a temporary queue. We publish three jobs, process them in parallel, save the results, and acknowledge each message. This example runs if `liteq`, `mirai`, and `jsonlite` are installed:

```{r, eval = requireNamespace("liteq", quietly = TRUE) && requireNamespace("mirai", quietly = TRUE) && requireNamespace("jsonlite", quietly = TRUE)}
# 1. Create a temporary queue and publish 3 jobs
db <- tempfile(fileext = ".sqlite")
q <- liteq::ensure_queue("jobs", db = db)
for (x in 1:3) {
    liteq::publish(q, title = paste0("job-", x),
                   message = jsonlite::toJSON(list(x = x), auto_unbox = TRUE))
}

# 2. Build the pipeline: process payloads on 2 parallel workers
mirai::daemons(2)

results <- list()
pipeline <- make_liteq_source(q, run_forever = FALSE) |>
    pump(function(payload) payload$x * 10,
         backend = mirai_backend(), max_workers = 2)

# 3. Drain: save successes; ack/nack happens automatically via lifecycle callbacks
pump_drain(pipeline, verbose = FALSE, handle_fn = function(id, data, ok) {
    if (ok && !inherits(data, "error")) {
        results[[as.character(id)]] <<- data
    }
})

mirai::daemons(0)

str(results)
```

Each job's payload `x` is multiplied by 10. The `liteq` message never leaves the main process - only the numeric payload is sent to a worker - and `ack()`/`nack()` are issued automatically via the item lifecycle callbacks once the result comes back.

## Step 3: The production daemon

A real daemon never stops on its own. Use `make_liteq_source(..., run_forever = TRUE)` (the default) and wrap the drain in `on.exit()` so workers always shut down on interrupt or error. Because `pump_drain()` streams each item and frees its memory, the process stays bounded even over millions of jobs:

```{r, eval = FALSE}
run_daemon <- function(db_path, queue_name = "jobs", n_workers = 4) {
    q <- liteq::ensure_queue(queue_name, db = db_path)

    save_result <- function(task_id, result) {
        # Persist to your database here
        message("Task ", task_id, " -> ", result)
    }

    mirai::daemons(n_workers)
    on.exit(mirai::daemons(0), add = TRUE)   # always release workers

    pipeline <- make_liteq_source(q) |>
        pump(function(payload) payload$x * 10,
             backend = mirai_backend(), max_workers = n_workers, buffer_size = n_workers)

    message("Daemon started. Polling queue: ", queue_name)
    pump_drain(
        pipeline,
        handle_fn = function(id, data, ok) {
            if (ok && !inherits(data, "error")) {
                save_result(id, data)
            }
        },
        sleep_ms = 100   # backoff when the queue is empty
    )
}
```

**Stopping the daemon.** Because the loop is infinite, stop it with an interrupt (`Ctrl-C` / `SIGINT`); the `on.exit()` hook then shuts the `mirai` workers down cleanly. For programmatic shutdown, publish a sentinel "poison-pill" message whose handler flips a flag that your `done_fn` checks.

## Advantages of this architecture

1. **Bounded concurrency and memory.** `max_workers` caps how many jobs run at once, protecting memory and database connections. Adding `buffer_size` applies buffer backpressure so the daemon stops pulling from the queue when result persistence is the bottleneck (see the backpressure section of the [main vignette](siphon.html)).
2. **Crash resilience.** A message is removed only after `liteq::ack()`. If the daemon crashes mid-task, the un-acked message is released by SQLite's lock mechanism and another worker retries it.
3. **Responsive UI.** A Shiny or Plumber front end only calls `liteq::publish()` to enqueue work, which returns instantly. It can then poll the database asynchronously (e.g. with a `later` loop or reactive timer) to check when the job's result has been saved.
